---
title: "Urban Planning and Ecology: The health outcomes of poor urban planning"
author: "Ricardo Gutierrez-Vallejo"
date: "February 10, 2020"
bibliography: references.bib
csl: ecology.csl
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Urban planning plays an important role in establishing ecosystem structure for local organisms and can even have cascading effects on human lifestyles. Urban planners utilize their knowledge of the environment and infrastrcture to design how cities can be organized to utlimately benefit the individuals that interact with these structures on a daily basis. Ecologist for some time have attempted to integrate theory into the field of urban planning, but not much progress has been made to date. Niemelä describes a three step process for successful integration, which includes a more methodical approach to urban planning resulting in improved conservation efforts [@niemela1999ecology]. By doing something as simple as going through a checklist, local environments can produce benefits for wildlife and humans. The benefits of these techniques span from improving human lifestyle, these also benefit global health in the form of reducing various high density areas with pollutants. To further demonstrate the effect of air toxics, a comparison performed across six U.S. cities depicted that areas with higher levels of toxics can often lead to an increased succeptibility to cardiovascular events [@suwa2002particulate].

Aside from conservation and global health, urban planning also plays an important role in terms of environmental health of a region. Disorganized urban planning can result in a combination of heavy air polluting facilties in close proximity of one another as demonstrated above. This in turn can have cascading effects on individuals living in the area. Due to this, individuals living in areas of cheap and accesible land for waste facilities often endure consequences of pollution. In a paper on environmental justice, Paul Mohai notes that there are socioeconomic effects of poor urban planning, indicating that some minority commmunities were often common sites for new waste facilities [@mohai2009environmental]. These social concerns led to many questions about the environmental conditions in which some minority communities have been forced into. Urban planning could be a potential route to resolving these potential racial issues, but no research has concretely demonstrated that poorer or disadvatnaged communities live in areas with unhealthy environmental conditions. A jornal article published a study using zip codes as areas of interests and demonstrated that race proved to be a critical variable when determining the location of waste facilities [@cutter1995race]. Along with this, researchers have often found that either race or socioeconomic status plays a role in the health and cleanliness of a community. 

One of the many pollutants in disadvantaged communities is the presence of chemical toxics generated by waste facillities, factories, vehicles, and other pollutants. A study of roadside metal contaminants performed in Shanghai, China identified that some toxic metals have potential to become "chemical time bombs" which could alter the urban ecosytem structure and eventually effect human health and safety [@shi2008potentially]. Occurences similar to this have become common place around the world regardless of socioeconomic status. The prominent issue stemming from this is that socioeconomically disadvatnaged individuals have a much harder time avoiding the negaive effects of pollution. Researchers have argued that through the accumulation of property and governmental influence, privelged Americans have constructed a heirarchy of status based on race and class [@hurley1995environmental]. These barriers to economic assension have enhanced the issue of ecological deterioration as the disadvantaged become more disadvantaged. 

To futher investigate the relationship between race and economic status and environmental health, I proposed a study involving the quantification of environmental polutants in the Marina, Tenderloin, and Richmond districts of San Francisco. In this report, I specifcally looked at the number of chemicals, classified as toxic, and whether these play a role in the decreased presence of green life density in selected areas of San Francisco. 

# Methods

Add text here.

## Site Locations and Descriptions

Add text here.

## Field Sampling Design

Add text here.

## Data Analysis and Statistics

Add text here.

# Results

## Subsections are ok in the results section too

Add a number of code chunks in the Results section. These should read in, subset and plot the data as needed (no need to save any figures to pdf, since they will be put into the rendered document when you click 'knit'), and, for any hypotheses that you want to test, an appropriate statistical test. 

```{r load-packages}
# If you add any additional packages here, make sure they are
# also listed in the DESCRIPTION file
library("dplyr")
library("tidyr")
library("ggplot2")
library("readr")
```

# Discussion

# Sources Cited

